Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography

### 2023-09-19 manuscript/wp4.tex
ms 19/09

Examples from other languages Irish orthography has special features to handle eclipsis. For example, when the word /pɑʃti/ (children) is eclipsed, it becomes /bɑʃti/ (the initial consonant is vocalized), but the spelling system retains the letter <p> \textit{in addition to} adding a <b>: <páistí> becomes <bpáistí> under eclipsis. This allows the reader quick access to the root word from which the eclipsed form is derived, while also showing the reader how to pronounce the eclipsed form. Moreover, in some cases, it results in ambiguity avoidance---in this case, if the <p> were not retained, the spelling <báistí> would be confusable with the word for rain.

@@

Participants were taught words for nine ``alien'' objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}. The alien words had a spoken and written form composed of a stem and suffix. The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus). These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent. Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time. Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the spelling of the suffix could potentially take on different forms to mark color.

@@

The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, reveling the proportion of the 10 chains that fall into each typological category at each generation. In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2. The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy. In Transmission~+ Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common. There is also a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings (i.e., K9, M1, M7, and Q7). For a statistical comparison across conditions, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
	\begin{equation}
	\begin{aligned}
	   t_i & \sim \mathrm{Multinomial}(10, \theta_i) \\
	   \theta_i & \sim \mathrm{Dirichlet}([1, 1, 1, 1]),
	\end{aligned}
	\end{equation}
and compute the posterior difference in $\theta_i$ (the estimated proportion of chains falling into each category) across conditions: $\Delta(\theta_i) = \theta_{i_\mathrm{comm}} - \theta_{i_\mathrm{trans}}$. The means and 90\% HDIs of these posterior differences are plotted in the bottom panels of Fig.~\ref{typ_dist_dif}. In general, we find no clear difference in the number of expressive systems between the two conditions; however, we do find that redundant systems are less common in Transmission~+ Communication (in most generations), while holistic systems tend to be a little more common (most notably at Generation~7). Although there is not much evidence to suggest that Transmission~+ Communication strongly favors more informative systems (i.e., the expressive and holistic categories), it is interesting to note that the communicative condition did seem to disfavor the inexpressive redundant systems.

@@

Finally, cost may be converted into the expected probability of communicative success with the equation $p=2^{-c}$. If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-\alpha} = 40.8$\% (90\% HDI: [35.1, 46.3]) in Transmission-only and 50.6\% (90\% HDI: [45.9, 54.9]) in Transmission~+ Communication, representing an advantage of around 10 percentage points for the communicative condition. Overall, then, while the presence of the communicative pressure does appear to make a meaningful difference to the informativeness of the emergent systems, the levels of informativeness that were achieved were not especially impressive considering that an expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use, while also being fairly easy to learn.

@@

So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. Faced with variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expulsion}, \textit{transmission}, and \textit{station}). As we will see in the following experiment, it is indeed possible for participants to learn expressive systems, even under full homophony; the lack of such outcomes in the present experiment suggests a reluctance to defy the spoken language.

@@

Many of the chains did, however, eschew informativeness entirely in favor of greater transparency, and the inevitable long-term consequence for all chains appears to be degeneracy. This is to be expected under Transmission-only, where the systems are adapting under learnability pressure, but is somewhat surprising in Transmission~+ Communication. Our findings ultimately suggest that, in the long term, (alphabetic) orthographic systems favor the faithful encoding of speech over the useful encoding of meaning, although there may exist short windows of time during which informative heterography can resist the spoken language. This points to a preference that readers might possess---those accustomed to an alphabetic writing system, at least---in which the primary goal of writing is seen as representing speech, not representing meaning directly. Interestingly, although participants were quite resistant to encoding into writing something that is not encoded in speech, they were---at the same time---content to conserve spelling forms that were internally inconsistent and unusual. Tradition has a powerful

@@

\begin{itemize}

	\item Summary of theoretical idea... hard to test in the real world, but we can experimentally simulate the cultural evolution of spelling

	\item Recapitulation of main findings

	\item What can this tell us about real-world processes?

	\item Cannot directly compare experiments because it's easier to maintain than create. However, this is also true of the real world, pointing to a general advantage of conservation over differentiation.

	\item The fact that the results are less "significant" in conservation does not mean that conservation is a worse model; it means that, under the conservation mechanism, communicative pressure is less important. This might suggest an accident-of-history explanation: Learning alone (i.e., habitual, mindless replication) can explain the emergence of informative orthography. Informative orthography does not arise through adaption, but may instead be the result of mindlessly replicated historic residue following sound change.

	\item Limitations

	\item Future directions. Remove the spoken forms from the test phase to simulate a silent-reader, literate society. This might make a dramatic difference to the results.

	\item Conclusion: perhaps there are potentially some benefits to heterography.

\end{itemize}

@@

>>>>

Irish orthography has special features to handle the initial consonant mutations that occur extensively across the Celtic languages. For example, under certain grammatical contexts, a word such as /pɑʃti/ (children) will be transformed into /bɑʃti/, a phenomenon known as eclipsis. The writing system, however, does not simply replace the <p> with a <b>, as one might predict; instead, both letters are included, allowing the orthography to, at~once, show meaning and phonetic form: the eclipsed spelling of <páistí> is <bpáistí>. Moreover, in some cases, the eclipsed spellings result in ambiguity avoidance---in this case, if the <p> were not retained, the hypothetical spelling <báistí> would be confusable with the genitive form of the word for rain.

Even the most orthographically transparent of languages have at least some techniques for disambiguating homophones in writing. Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}). Many other languages employ various diacritics for the same purpose, such as the acute accent in Danish---<en> (\textit{a}) vs.\@ <én> (\textit{one})---or the circumflex in Serbo-Croatian---<da> (\textit{yes}) vs.\@ <dâ> (\textit{gives}).

@@

Participants were taught words for nine ``alien'' objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}. The alien words had a spoken and written form composed of a stem and suffix. The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus). These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent. Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time. Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color.

@@

The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, reveling the proportion of the 10 chains that fall into each typological category at each generation. In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2. The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy. In Transmission~+ Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common. There is also a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings (i.e., K9, M1, M7, and Q7). To quantify our uncertainty about these distributions and compare conditions statistically, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
	\begin{equation}
	\begin{aligned}
	   t_i & \sim \mathrm{Multinomial}(10, \theta_i) \\
	   \theta_i & \sim \mathrm{Dirichlet}([1, 1, 1, 1]),
	\end{aligned}
	\end{equation}
and compute the posterior difference in $\theta_i$ (the estimated proportion of chains falling into each category) across conditions: $\Delta(\theta_i) = \theta_{i_\mathrm{comm}} - \theta_{i_\mathrm{trans}}$. The means and 90\% HDIs of these posterior differences are plotted in the bottom panels of Fig.~\ref{typ_dist_dif}. In general, we find no clear difference in the number of expressive systems between the two conditions; however, we do find that redundant systems are less common in Transmission~+ Communication (in most generations), while holistic systems tend to be a little more common (most notably at Generation~7). Although there is not much evidence to suggest that Transmission~+ Communication strongly favors more informative systems (i.e., the expressive and holistic categories), it is interesting to note that the communicative condition did seem to disfavor the inexpressive redundant systems.

@@

Finally, cost may be converted into the expected probability of communicative success\footnote{Note that we can compute this quantity regardless of which condition a participant was in. By expected probability of communicative success, we mean the level of success a pair of interlocutors could hypothetically achieve using a given system, regardless of whether that system emerged in the Transmission-only or Transmission~+ Communication condition. This also assumes that the hypothetical interlocutors have a perfect understanding of the same shared system.} with the equation $p=2^{-c}$. If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-\alpha} = 40.8$\% (90\% HDI: [35.1, 46.3]) in Transmission-only and 50.6\% (90\% HDI: [45.9, 54.9]) in Transmission~+ Communication, representing an advantage of around 10 percentage points for the communicative condition. Overall, then, while the presence of the communicative pressure does appear to make a meaningful difference to the informativeness of the emergent systems, the levels of informativeness that were achieved were not especially impressive considering that a degenerate system would result in a communicative success rate of~33.3\% (i.e., chance-level performance), while an expressive system would result in a~100\% success rate, while also being fairly easy to learn.

@@

So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. Faced with variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expulsion}, \textit{transmission}, and \textit{station}).

@@

Many of the chains did, however, eschew informativeness entirely in favor of greater transparency, and the inevitable long-term consequence for all chains appears to be degeneracy. This is to be expected under Transmission-only, where the systems are adapting under learnability pressure, but is somewhat surprising in Transmission~+ Communication. Our findings ultimately suggest that, in the long term, (alphabetic) orthographic systems favor the faithful encoding of speech over the useful encoding of meaning, although there may exist short windows of time during which informative heterography can resist the spoken language. This points to a preference that readers might possess---those accustomed to an alphabetic writing system, at least---in which the primary goal of writing is seen as representing speech, not representing meaning directly. Interestingly, although participants were quite resistant to encoding into writing something that is not encoded in speech, they were---at the same time---content to conserve spelling forms that were internally inconsistent and unusual. Tradition has a powerful hold over writing systems.

@@

Writing---particularly writing in alphabetic scripts---is usually viewed as a visual representation of speech: Each visual letter corresponds to a sound. However, writing systems are known to diverge from their spoken counterparts in various ways. The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech. It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality. In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality. In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language, if that divergence confers some benefit in reading.

We tested this hypothesis by focusing on the particular case of heterographic homophones---words that sound the same but which are spelled differently. Heterographic homophones permit the written language to be more informative than the spoken language. The written forms <knight> and <night>, for example, convey a distinction in meaning that cannot be conveyed in speech without supplying additional information, either in the form of context or paraphrasis. In English, at~least, which has a tendency to preserve etymology in spelling, the heterographic homophones have typically been viewed as accidents of history: \textit{right}, \textit{rite}, and \textit{write} are spelled differently because their spellings have been inherited from the conventions of the languages of origin (Old English \textit{riht}, Latin \textit{ritus}, and Old English \textit{writan}). Nevertheless, examples of explicit efforts to differentiate the spelling of homophones do exist, such as x, y, z, and, moreover, the etymological cases of heterography may, in part, owe their own success to the differentiation they inadvertently provide.

We put these ideas to the test by experimentally simulating the cultural evolution of orthography. In Experiment~1, we focused on the differentiation mechanism: Might variant spellings be used to differentiate meanings that are otherwise identical in speech? If so, we would expect levels of spelling differentiation to be greater when there is greater pressure for disambiguation, which we induced through the addition of a communication game. Importantly, the initial randomly generated orthographic systems that we used to seed the transmission chains contained high variation---that is, multiple ways of spelling the same sound. We included this variation because, for the differentiation mechanism to be viable, the writing system has to be receptive to spelling variation. A writing system that does not permit a one-to-many phoneme-to-grapheme mapping would not be capable of differentiating homophonous words. Only when variation is permitted and available can variant spellings be conditioned on meaning.

Although we found evidence to suggest that the written forms were slightly more informative than the corresponding spoken forms, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but conditioned on shape rather than color. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems. Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors: First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech. Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express. Secondly, even when participants appreciate the need to differentiate the written forms to be successful in the task, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome. It is notable, for example, that in the communication games, many attempts to differentiate with English color letters were not reciprocated.

In Experiment~2, we tested a different mechanism by which orthographies may end up possessing additional informativeness beyond that of the spoken form of the language: conservation. Under this mechanism, an expressive orthography does not emerge to fulfill a role but rather it is simply maintained from an earlier form of the spoken language that was expressive. This parallels what we know about many of the heterographic homophones in English, which arose as byproducts of either the preservation of etymology or phonological changes that were never assimilated into the written forms \parencite{Berg:2021}. That is not to say, necessarily, that this form of heterography is entirely an accident of history; rather, such spellings may, at least in part, have been preserved precisely because of the informativeness they provide in reading. If correct, this would predict that expressive orthography should be preserved preferentially under communicative pressure.

Our findings showed that informative heterography may be conserved more frequently and for longer periods under communicative pressure for disambiguation. There are two important caveats, however. First, we found that cultural transmission alone---that is, blind learning and reproduction---will result in at~least some conservation, not only in form but also in the conditioning of form on meaning. Cases such as Chain~D correspond to the ``accident of history'' explanation: Expressive orthography is preserved not because it serves any useful purpose (recall that in Transmission-only there is no functional need for the language to be informative), but because participants are simply reproducing what they learned, and what they learned has not (yet) placed a significant enough burden on learning for simplification to kick~in. The additional level of conservation that occurs in Transmission~+ Communication corresponds to the exaptation/repurposing explanation; that is, expressive orthography that originally served one purpose (representing speech) is maintained for a new purpose (representing meaning directly). The second caveat is that, in the long term, it appears that transparency will ultimately win the day, even under communicative pressure. Chain~O, for example, went from expressive to degenerate in two generations under full homophony pressure, and it seems likely that all chains will ultimately undergo the same transformation eventually. Informative heterography that arises through conservation is but a temporary oasis on the march toward transparency.

It is important to note, at this point, that the two experiments cannot be compared directly, although we made every effect to keep the two as close as possible. Fundamentally, participants---or more generally, the evolutionary systems---are being asked to do something quite different across the two experiments: create in Experiment~1 and maintain in Experiment~2. The demands of these two tasks are different, and one task or the other may be better suited to our experimental paradigm. Similarly, we cannot draw direct conclusions from the relative strength of the results across the two experiments. For example, although the difference between conditions was perhaps a little more dramatic in Experiment~1, this does not imply that the differentiation model is a better candidate mechanism; it suggests, rather, that communicative pressure is more essential for the differentiation mechanism to operate. Nevertheless, our experiments do serve to highlight the comparative difficulties involved in differentiation vs.\@ conservation. For differentiation to operate, numerous hurdles must be overcome: participants must recognize the problem they are faced with, they must be willing to introduce and tolerate spelling variation, they must able to align with an interlocutor separated in time and space, and they must be willing to rebel against their input. Conserving an expressive orthography is, by comparison, a relatively simple affair. If conservation is a simpler (and therefore, by Occam's razor, more plausible) mechanism in our experimental setup, then presumably the same is true of the real world.

Of course, all of these conclusions must be interpreted within the limitations of our studies; our experiments are, after all, highly simplified simulacra of the relevant real-world processes. Other than the general scaling-down of the x to an experimentally tractable

Communication game not a good model of written communication. link to texting which is written but near-synchronous.

color morphs a bit weird

do not consider top-down authority

unclear whether disambiguation matters in the real world. context usually makes things clear.

Nevertheless, even very small biases can be amplified by cultural transmission.

We foresee two interesting directions in which this work might be taken in the future. The first is a relatively simple extension of the present experiments. By removing the spoken forms from the test/communication~game phase of the experiment, we can simulate something closer to a society of silent readers; that is, a society wherein the spoken form is only salient during learning and less salient once the users switch to using the language primarily through the written modality. This minor change might have a dramatic effect on the results: Participants may be less inclined to simply transcribe how a word sounds and more inclined to differentiate and conserve. Such an experiment is not without pitfalls, however. Notably, our reason for providing the spoken form at the test stage was to be sure that participants were aware of the homophony; our experiment aims to test whether participants can overcome homophony, so it was vital to be sure that participants were indeed aware of it. Thus, this change to the experimental design might be open to concerns that, at the test stage, participants might simply forget everything about the pronunciation, such that the experiment becomes equivalent to prior iterated learning studies---that is, it becomes focused on the emergence of linguistic structure in general, not orthographic structure that runs contrary to the spoken language.

A second, more general direction in which this work could be taken would be to explore how well the differentiation and conservation models of heterography fit with historical data on language change. For example, by tracing the development of the heterographic homophones in English, we might be able to determine which of the models offer a better fit to the data, or indeed classify different cases as being derived through the differentiation or conservation mechanism. Perhaps the best way to proceed here would be to develop formal models of these two mechanisms and test them against diachronic data on the pronunciation and spelling of words that ultimately become heterographic.

As a final note, heterography is often derided as a source of unnecessary complexity that language learners must contend with. However, our results show that, when the communicative need for disambiguation is strong enough, language users may prefer to tradeoff some simplicity for greater informativeness in the writing system. Perhaps, then, efforts to reform and simplify spelling may, to some extent, be a little hasty, and more recognition should be given to the benefits that opaque orthographies provide in the context of reading.

@@

\clearpage

\clearpage

\clearpage
